Comparing and combining network-guided biomarker discovery to discover genetic susceptibility mechanisms to breast cancer on the GENESIS study

### 2020-08-20 manuscript/plos/original.tex
prepared review environment

% Template for PLoS
% Version 3.5 March 2018
%
% % % % % % % % % % % % % % % % % % % % % %
%
% -- IMPORTANT NOTE
%
% This template contains comments intended 
% to minimize problems and delays during our production 
% process. Please follow the template instructions
% whenever possible.
%
% % % % % % % % % % % % % % % % % % % % % % % 
%
% Once your paper is accepted for publication, 
% PLEASE REMOVE ALL TRACKED CHANGES in this file 
% and leave only the final text of your manuscript. 
% PLOS recommends the use of latexdiff to track changes during review, as this will help to maintain a clean tex file.
% Visit https://www.ctan.org/pkg/latexdiff?lang=en for info or contact us at latex@plos.org.
%
%
% There are no restrictions on package use within the LaTeX files except that 
% no packages listed in the template may be deleted.
%
% Please do not include colors or graphics in the text.
%
% The manuscript LaTeX source should be contained within a single file (do not use \input, \externaldocument, or similar commands).
%
% % % % % % % % % % % % % % % % % % % % % % %
%
% -- FIGURES AND TABLES
%
% Please include tables/figure captions directly after the paragraph where they are first cited in the text.
%
% DO NOT INCLUDE GRAPHICS IN YOUR MANUSCRIPT
% - Figures should be uploaded separately from your manuscript file. 
% - Figures generated using LaTeX should be extracted and removed from the PDF before submission. 
% - Figures containing multiple panels/subfigures must be combined into one image file before submission.
% For figure citations, please use "Fig" instead of "Figure".
% See http://journals.plos.org/plosone/s/figures for PLOS figure guidelines.
%
% Tables should be cell-based and may not contain:
% - spacing/line breaks within cells to alter layout or alignment
% - do not nest tabular environments (no tabular environments within tabular environments)
% - no graphics or colored text (cell background color/shading OK)
% See http://journals.plos.org/plosone/s/tables for table guidelines.
%
% For tables that exceed the width of the text column, use the adjustwidth environment as illustrated in the example table in text below.
%
% % % % % % % % % % % % % % % % % % % % % % % %
%
% -- EQUATIONS, MATH SYMBOLS, SUBSCRIPTS, AND SUPERSCRIPTS
%
% IMPORTANT
% Below are a few tips to help format your equations and other special characters according to our specifications. For more tips to help reduce the possibility of formatting errors during conversion, please see our LaTeX guidelines at http://journals.plos.org/plosone/s/latex
%
% For inline equations, please be sure to include all portions of an equation in the math environment.  For example, x$^2$ is incorrect; this should be formatted as $x^2$ (or $\mathrm{x}^2$ if the romanized font is desired).
%
% Do not include text that is not math in the math environment. For example, CO2 should be written as CO\textsubscript{2} instead of CO$_2$.
%
% Please add line breaks to long display equations when possible in order to fit size of the column. 
%
% For inline equations, please do not include punctuation (commas, etc) within the math environment unless this is part of the equation.
%
% When adding superscript or subscripts outside of brackets/braces, please group using {}.  For example, change "[U(D,E,\gamma)]^2" to "{[U(D,E,\gamma)]}^2". 
%
% Do not use \cal for caligraphic font.  Instead, use \mathcal{}
%
% % % % % % % % % % % % % % % % % % % % % % % % 
%
% Please contact latex@plos.org with any questions.
%
% % % % % % % % % % % % % % % % % % % % % % % %

@@

@@

\newcommand{\lorem}{{\bf LOREM}}
\newcommand{\ipsum}{{\bf IPSUM}}

@@

% Title must be 250 characters or less.
\begin{flushleft}
{\Large
\textbf\newline{Title of submission to PLOS journals} % Please use "sentence case" for title and headings (capitalize only the first word in a title (or heading), the first word in a subtitle (or subheading), and any proper nouns).
}
\newline
% Insert author names, affiliations and corresponding author email (do not include titles, positions, or degrees).
\\
Name1 Surname\textsuperscript{1,2\Yinyang},
Name2 Surname\textsuperscript{2\Yinyang},
Name3 Surname\textsuperscript{2,3\textcurrency},
Name4 Surname\textsuperscript{2},
Name5 Surname\textsuperscript{2\ddag},
Name6 Surname\textsuperscript{2\ddag},
Name7 Surname\textsuperscript{1,2,3*},
with the Lorem Ipsum Consortium\textsuperscript{\textpilcrow}
\\
\bigskip
\textbf{1} Affiliation Dept/Program/Center, Institution Name, City, State, Country
\\
\textbf{2} Affiliation Dept/Program/Center, Institution Name, City, State, Country
\\
\textbf{3} Affiliation Dept/Program/Center, Institution Name, City, State, Country
\\
\bigskip

% Insert additional author notes using the symbols described below. Insert symbol callouts after author names as necessary.
% 
% Remove or comment out the author notes below if they aren't used.
%
% Primary Equal Contribution Note
\Yinyang These authors contributed equally to this work.

% Additional Equal Contribution Note
% Also use this double-dagger symbol for special authorship notes, such as senior authorship.
\ddag These authors also contributed equally to this work.

% Current address notes
\textcurrency Current Address: Dept/Program/Center, Institution Name, City, State, Country % change symbol to "\textcurrency a" if more than one current address note
% \textcurrency b Insert second current address 
% \textcurrency c Insert third current address

% Deceased author note
\dag Deceased

@@

% Use the asterisk to denote corresponding authorship and provide email address in note below.
* correspondingauthor@institute.edu

\end{flushleft}
% Please keep the abstract below 300 words
\section*{Abstract}
Lorem ipsum dolor sit amet, consectetur adipiscing elit. Curabitur eget porta erat. Morbi consectetur est vel gravida pretium. Suspendisse ut dui eu ante cursus gravida non sed sem. Nullam sapien tellus, commodo id velit id, eleifend volutpat quam. Phasellus mauris velit, dapibus finibus elementum vel, pulvinar non tellus. Nunc pellentesque pretium diam, quis maximus dolor faucibus id. Nunc convallis sodales ante, ut ullamcorper est egestas vitae. Nam sit amet enim ultrices, ultrices elit pulvinar, volutpat risus.

% Please keep the Author Summary between 150 and 200 words
% Use first person. PLOS ONE authors please skip this step. 
% Author Summary not valid for PLOS ONE submissions.   
\section*{Author summary}
Lorem ipsum dolor sit amet, consectetur adipiscing elit. Curabitur eget porta erat. Morbi consectetur est vel gravida pretium. Suspendisse ut dui eu ante cursus gravida non sed sem. Nullam sapien tellus, commodo id velit id, eleifend volutpat quam. Phasellus mauris velit, dapibus finibus elementum vel, pulvinar non tellus. Nunc pellentesque pretium diam, quis maximus dolor faucibus id. Nunc convallis sodales ante, ut ullamcorper est egestas vitae. Nam sit amet enim ultrices, ultrices elit pulvinar, volutpat risus.

@@

\section{Introduction}

@@

\section{Methods}
\subsection{GENESIS}

>>>>

% Template for PLoS
% Version 3.5 March 2018
%
% % % % % % % % % % % % % % % % % % % % % %
%
% -- IMPORTANT NOTE
%
% This template contains comments intended 
% to minimize problems and delays during our production 
% process. Please follow the template instructions
% whenever possible.
%
% % % % % % % % % % % % % % % % % % % % % % % 
%
% Once your paper is accepted for publication, 
% PLEASE REMOVE ALL TRACKED CHANGES in this file 
% and leave only the final text of your manuscript. 
% PLOS recommends the use of latexdiff to track changes during review, as this will help to maintain a clean tex file.
% Visit https://www.ctan.org/pkg/latexdiff?lang=en for info or contact us at latex@plos.org.
%
%
% There are no restrictions on package use within the LaTeX files except that 
% no packages listed in the template may be deleted.
%
% Please do not include colors or graphics in the text.
%
% The manuscript LaTeX source should be contained within a single file (do not use \input, \externaldocument, or similar commands).
%
% % % % % % % % % % % % % % % % % % % % % % %
%
% -- FIGURES AND TABLES
%
% Please include tables/figure captions directly after the paragraph where they are first cited in the text.
%
% DO NOT INCLUDE GRAPHICS IN YOUR MANUSCRIPT
% - Figures should be uploaded separately from your manuscript file. 
% - Figures generated using LaTeX should be extracted and removed from the PDF before submission. 
% - Figures containing multiple panels/subfigures must be combined into one image file before submission.
% For figure citations, please use "Fig" instead of "Figure".
% See http://journals.plos.org/plosone/s/figures for PLOS figure guidelines.
%
% Tables should be cell-based and may not contain:
% - spacing/line breaks within cells to alter layout or alignment
% - do not nest tabular environments (no tabular environments within tabular environments)
% - no graphics or colored text (cell background color/shading OK)
% See http://journals.plos.org/plosone/s/tables for table guidelines.
%
% For tables that exceed the width of the text column, use the adjustwidth environment as illustrated in the example table in text below.
%
% % % % % % % % % % % % % % % % % % % % % % % %
%
% -- eqnarrayS, MATH SYMBOLS, SUBSCRIPTS, AND SUPERSCRIPTS
%
% IMPORTANT
% Below are a few tips to help format your eqnarrays and other special characters according to our specifications. For more tips to help reduce the possibility of formatting errors during conversion, please see our LaTeX guidelines at http://journals.plos.org/plosone/s/latex
%
% For inline eqnarrays, please be sure to include all portions of an eqnarray in the math environment.  For example, x$^2$ is incorrect; this should be formatted as $x^2$ (or $\mathrm{x}^2$ if the romanized font is desired).
%
% Do not include text that is not math in the math environment. For example, CO2 should be written as CO\textsubscript{2} instead of CO$_2$.
%
% Please add line breaks to long display eqnarrays when possible in order to fit size of the column. 
%
% For inline eqnarrays, please do not include punctuation (commas, etc) within the math environment unless this is part of the eqnarray.
%
% When adding superscript or subscripts outside of brackets/braces, please group using {}.  For example, change "[U(D,E,\gamma)]^2" to "{[U(D,E,\gamma)]}^2". 
%
% Do not use \cal for caligraphic font.  Instead, use \mathcal{}
%
% % % % % % % % % % % % % % % % % % % % % % % % 
%
% Please contact latex@plos.org with any questions.
%
% % % % % % % % % % % % % % % % % % % % % % % %

@@

% author's packages
\usepackage[shortcuts]{extdash}
\usepackage{lmodern} % Chloe: better font rendering

@@

\newcommand{\lorem}{{\bf LOREM}}
\newcommand{\ipsum}{{\bf IPSUM}}
\newcommand{\mean}[1]{$\overline{\mbox{#1}}$}
\newcommand{\median}[1]{$\hat{\mbox{#1}}$}

@@

% Title must be 250 characters or less.
\begin{flushleft}
{\Large
\textbf\newline{Comparing and combining network-guided biomarker discovery to discover genetic susceptibility mechanisms to breast cancer on the GENESIS study} % Please use "sentence case" for title and headings (capitalize only the first word in a title (or heading), the first word in a subtitle (or subheading), and any proper nouns).
}
\newline
% Insert author names, affiliations and corresponding author email (do not include titles, positions, or degrees).
\\
Héctor Climente-González\textsuperscript{1,2,3,4*}, 
Christine Lonjou\textsuperscript{1,2,3}, 
Fabienne Lesueur\textsuperscript{1,2,3},
Dominique Stoppa-Lyonnet\textsuperscript{5,6,7\textcurrency}, 
Nadine Andrieu\textsuperscript{1,2,3}, 
Chloé-Agathe Azencott\textsuperscript{3,1,2},
with the GENESIS study group\textsuperscript{\textpilcrow}
\\
\bigskip
\textbf{1} Institut Curie, PSL Research University, F-75005 Paris, France;\\
\textbf{2} INSERM, U900, F-75005 Paris, France;\\
\textbf{3} MINES ParisTech, PSL Research University, CBIO-Centre for Computational Biology, F-75006 Paris, France;\\
\textbf{4} RIKEN Center for Advanced Intelligence Project (AIP), Tokyo, Japan; \\
\textbf{5} Service de Génétique, Institut Curie, F-75005 Paris, France;\\
\textbf{6} INSERM, U830, F-75005 Paris, France;\\
\textbf{7} Université Paris Descartes.\\
\bigskip

% Insert additional author notes using the symbols described below. Insert symbol callouts after author names as necessary.
% 
% Remove or comment out the author notes below if they aren't used.
%
% Primary Equal Contribution Note
% \Yinyang These authors contributed equally to this work.

% Additional Equal Contribution Note
% Also use this double-dagger symbol for special authorship notes, such as senior authorship.
% \ddag These authors also contributed equally to this work.

% Current address notes
\textcurrency For the GENESIS study group % change symbol to "\textcurrency a" if more than one current address note
% \textcurrency b Insert second current address 
% \textcurrency c Insert third current address

% Deceased author note
% \dag Deceased

@@

% Use the asterisk to denote corresponding authorship and provide email address in note below.
* hector.climente@riken.jp

\end{flushleft}
% Please keep the abstract below 300 words
\section*{Abstract}
Systems biology provides a comprehensive approach to biomarker discovery and biological hypothesis building. Indeed, it allows to jointly consider the statistical association between gene variation and a phenotype, and the biological context of each gene, represented as a network. In this work, we study six network methods which identify subnetworks with high association scores to a phenotype. Specifically, we examine their utility to discover new biomarkers for breast cancer susceptibility by interrogating a genome-wide association study (GWAS) focused on French women with a family history of breast cancer and tested negative for pathogenic variants in \emph{BRCA1} and \emph{BRCA2}. We perform an in-depth benchmarking of the methods with regards to size of the solution subnetwork, their utility as biomarkers, and the stability and the runtime of the methods. By trading statistical stringency for biological meaningfulness, most network methods give more compelling results than standard SNP- and gene-level analyses, recovering causal subnetworks tightly related to cancer susceptibility. For instance, we show a general alteration of the neighborhood of \emph{COPS5}, a gene related to multiple hallmarks of cancer. Importantly, we find a significantly large overlap between the genes in the solution networks and the genes significantly associated in the largest GWAS on susceptibility to breast cancer. Yet, network methods are notably unstable, producing different results when the input data changes slightly. To account for that, we produce a stable consensus subnetwork, formed by the most consistently selected genes. The stable consensus is composed of 68 genes, enriched in known breast cancer susceptibility genes (\emph{BLM}, \emph{CASP8}, \emph{CASP10}, \emph{DNAJC1}, \emph{FGFR2}, \emph{MRPS30}, and \emph{SLC4A7}, Fisher's exact test P\=/value~=~$3 \times 10^{-4}$) and occupying more central positions in the network than average. The network seems organized around \emph{CUL3}, encoding an ubiquitin ligase related protein that regulates the protein levels of several genes involved in cancer progression. In conclusion, this article shows the pertinence of network-based analyses to tackle known issues with GWAS, namely lack of statistical power and of interpretable solutions. Project-agnostic implementations of each of the network methods are available at \url{https://github.com/hclimente/gwas-tools} to facilitate their application to other GWAS datasets.

% Please keep the Author Summary between 150 and 200 words
% Use first person. PLOS ONE authors please skip this step. 
% Author Summary not valid for PLOS ONE submissions.   
\section*{Author summary}
In genome-wide association studies (GWAS), thousands of genomes are scanned to identify variants associated with a complex trait. Over the last 15 years, GWAS have advanced our understanding of the genetics of complex diseases, and in particular of hereditary cancers. Yet, they have led to an apparent paradox: the more we perform such studies, the more it seems that the entire genome is involved in every disease. An elegant explanation has been proposed with the omnigenic model: only a limited number of core genes are directly involved in the disease; but gene functions are deeply interrelated, so that many other genes are able to alter the function of the core genes. These interrelations are often modeled as networks, and multiple algorithms have been proposed to use these networks to identify the subset of core genes involved in a specific trait. In this study, we characterize six such network methods on GENESIS, a GWAS dataset for familial breast cancer in the French population. Combining these approaches allows us to identify potentially novel breast cancer susceptibility genes, and provides a mechanistic explanation for their role in the development of the disease. Our pipeline can easily be applied to other diseases.

@@

% Use "Eq" instead of "eqnarray" for eqnarray citations.
% For figure citations, please use "Fig" instead of "Figure".
% Place figure captions after the first paragraph in which they are cited.
\section{Introduction}

@@

% Results and Discussion can be combined.
\section{Results}

We conducted association analyses in the GENESIS dataset at both SNP and gene levels (Section~\ref{methods:node_score}). Two genomic regions have a P\=/value lower than the Bonferroni threshold on chromosomes 10 and 16 (\nameref{sfig:snp_gene_manhattan}A). The former overlaps with gene \emph{FGFR2}; the latter with \emph{CASC16}, and it is located near the protein-coding gene \emph{TOX3}. Variants in both \emph{FGFR2} and \emph{TOX3} have been repeatedly associated with breast cancer susceptibility in other case-control studies \cite{Michailidou2017}, \emph{BRCA1} and \emph{BRCA2} carrier studies \cite{Mulligan2011}, and in hereditary breast and ovarian cancer families negative for mutations in \emph{BRCA1} and \emph{BRCA2} \cite{rinella_genetic_2013}. In our studied population, only \emph{FGFR2} was significantly associated with breast cancer at the gene-level (\nameref{sfig:snp_gene_manhattan}B).

Closer examination reveals two regions (3p24 and 8q24) having low, albeit not genome-wide significant, P\=/values. Both of them have been associated to breast cancer susceptibility in the past \cite{brisbin_meta-analysis_2011,search_newly_2009}. We applied an L1-penalized logistic regression on the whole dataset (Section~\ref{methods:classifier}). It selected 100 SNPs, both from all aforementioned regions and new ones (\nameref{sfig:snp_gene_manhattan}C). Yet, it is unclear why those SNPs were selected, as emphasized by the high P\=/value of some of them, which further complicates the biological interpretation. Moreover, and in opposition to what would be expected under the omnigenic model, the genes to which these SNPs map to (Section~\ref{methods:node_score}) are not interconnected in the PPIN (Section~\ref{methods:networks}). In addition, the classification performance of the method is very low, and L1-penalized methods select only one of several correlated variables and are prone to instability, which further complicates interpretation. This motivates exploring network methods, which trade statistical significance for biological relevance to find susceptibility subnetworks. In fact, such methods provided comparably (poor) classification performance to L1-penalized logistic regression (Fig~\ref{fig:benchmark}B), while providing more interpretable solutions.

\subsection{Network methods successfully identify genes associated with breast cancer}
\label{results:separate_networks}

We applied six network methods to the GENESIS dataset (Section~\ref{methods:methods}). As none of the networks examined by LEAN was significant (BH adjusted P\=/value < 0.05), we obtained six solutions (Fig~\ref{fig:solution_overview}): one for each of the remaining four gene-based methods, one for SConES GI (which works at the SNP level), and the consensus. These solutions differ in many aspects, making it hard to draw joint conclusions. For starters, the overlap between the genes featured in each solution is quite small (Fig~\ref{fig:solution_overview}B). Another prominent difference is their size: the largest solution, produced by HotNet2, contains 440 genes, while heinz's contained only 4 genes. While SConES GI failed to recover any protein coding gene, by dealing with SNP networks it retrieved four subnetworks in intergenic regions, and another one overlapping an RNA gene (\emph{RNU6-420P}). Their topologies also differ, as measured by the median centrality (Table~\ref{tab:gene_solutions}) and the number of connected components (Fig~\ref{fig:solution_overview}D). Only two methods have more than one connected component: SConES, as described above, and HotNet2. HotNet2 produced 135 subnetworks, 115 of which have less than five genes. The second largest subnetwork (13 nodes) contains the two breast cancer susceptibility genes \emph{CASP8} and \emph{BLM}. Lastly, a pathway enrichment analysis (Section \ref{methods:pathway_enrichment}) also showed similarities and differences in the underlying mechanisms. It linked different parts of SigMod's solution network to four processes: protein translation (including mitochondrial), mRNA splicing, protein misfolding, and keratinization (BH adjusted P\=/values~<~0.03). Interestingly, dmGWAS solution is also related to protein misfolding (\emph{attenuation phase}, BH adjusted P\=/value~=~0.01). But, additionally, it includes submodules of proteins related to mitosis, DNA damage, and regulation of TP53 (BH adjusted P\=/values~<~0.05), which match previously known mechanisms of breast cancer susceptibility \cite{nielsen_hereditary_2016}. As with SigMod, the genes in HotNet2's solution are involved in mitochondrial translation (BH adjusted P\=/value~=~1.87 \texttimes{} 10\textsuperscript{-4}), but also in glycogen metabolism and transcription of nuclear receptors (BH adjusted P\=/value~<~0.04).

\begin{table}[!ht]
  \begin{adjustwidth}{-2.25in}{0in} % Comment out/remove adjustwidth environment if table fits in text column.
  \centering
  \caption{
  {\bf Summary statistics on the solutions of multiple network methods on the gene-gene interaction network. The first row contains the summary statistics on the whole network.}}
  \begin{tabular}{lrrrrr}
  {\bf Network } & {\bf \# genes } & {\bf \# edges } & {\bf \mean{Betweenness} } & {\bf \median{P}\textsubscript{gene} } & {\bf \(\rho\)\textsubscript{consensus} }\\
  \thickhline
  HINT HT & 13\,619 & 142\,541 & 16\,706 & 0.46 & 0.066\\
  \hline
  dmGWAS & 194 & 450 & 49\,115 & 0.19 & 0.41\\
  heinz & 4 & 3 & 113\,633 & 0.001 & 0.21\\
  HotNet2 & 440 & 374 & 7\,739 & 0.048 & 0.31\\
  LEAN & 0 & 0 & - & - & -\\
  SConES GI & 0 (1) & 0 & - & - & -\\
  SigMod & 142 & 249 & 92\,603 & 0.008 & 0.73\\
  \hline
  Consensus & 93 & 186 & 50\,737 & 0.006 & 1\\
  Stable consensus & 68 & 49 & 94\,854 & 0.005 & 0.54 \\
  \end{tabular}
  \begin{flushleft} \textbf{\emph{\# genes}}: number of genes selected out of those that are part of the PPIN; for SConES GI the total number of genes, including RNA genes, was added in parentheses. \textbf{\mean{Betweenness}}: mean betweenness of the selected genes in the PPIN. \textbf{\median{P}\textsubscript{gene}}: median P\=/value of the selected genes. \textbf{\(\rho\)\textsubscript{consensus}}: Pearson correlation between the subnetwork and the consensus network.
  \end{flushleft}
  \label{tab:gene_solutions}
  \end{adjustwidth}
\end{table}

\begin{figure}[!ht]
  \centering
  \includegraphics[height=\textwidth]{./figures/figure_1.png}
  \caption{{\bf Overview of the solutions produced by the different network methods (Section~\ref{methods:methods}) on the GENESIS dataset. } As LEAN did not produce any significant gene (BH adjusted P\=/value~<~0.05), it was excluded. Unless indicated otherwise, results refer to genes, except for SConES GI which are at the SNP-level. \textbf{(A)} Manhattan plots of SNPs/genes; in black, the method's solution. The Bonferroni threshold is indicated by a red line (2.54~\texttimes{}~10\textsuperscript{-7} for SNPs, 1.53~\texttimes{}~10\textsuperscript{-6} for genes). \textbf{(B)} Overlap between the genes selected by each of the methods, measured by Pearson correlation between indicator vectors. \textbf{(C)} Precision and recall of the evaluated methods with respect to Bonferroni-significant SNPs/genes in BCAC. For reference, we added a gray line with a slope of 1. \textbf{(D)} Solution networks produced by the different methods.}
  \label{fig:solution_overview}
\end{figure}

Despite their differences, there are additional common themes. All obtained solution subnetworks have lower association P\=/values than the whole PPIN (median P\=/value $\ll 0.46$, Table~\ref{tab:gene_solutions}), despite containing genes with higher P\=/values as well (Fig~\ref{fig:solution_overview}A). This exemplifies the trade-off between statistical significance and biological relevance. However, there are nuances between solutions in this regard: heinz strongly favored genes with lower P\=/values, while dmGWAS was less conservative (median P\=/values 0.0012 and 0.19, respectively); SConES tended to select whole LD-blocks; and HotNet2 and SigMod were less likely to select low scoring genes. Additionally, the solution subnetworks presented other desirable properties. First, five of them were enriched in known breast cancer susceptibility genes (consensus, dmGWAS, heinz, HotNet2, and SigMod, Fisher's exact test one-sided P\=/value < 0.03). Second, the genes in four solution subnetworks displayed on average a higher betweenness centrality than the rest of the genes, a difference that is significant in four solutions (consensus, dmGWAS, HotNet2, and SigMod, Wilcoxon rank-sum test P\=/value < 1.4 \texttimes{} 10\textsuperscript{-21}). This agrees with the notion that disease genes are more central than other non-essential genes \cite{pinero_uncovering_2016}, an observation that holds in breast cancer (one-tailed Wilcoxon rank-sum test P\=/value~=~2.64 \texttimes{} 10\textsuperscript{-5} when comparing the betweenness of known susceptibility genes versus the rest). Interestingly, SConES selected SNPs that are also more central than the average SNP (Supplementary table~\nameref{stab:snp_solutions}), suggesting that causal SNPs are also more central than non associated SNPs.

\subsection{A case study: the consensus network}
\label{results:consensus}

Despite the heterogeneity of the solutions, their shared properties suggest that each method captures different aspects of cancer susceptibility. Indeed, only 20 genes are common to more than two solutions (\nameref{sfig:consensus_stats}A), but encouragingly, the more methods selected a gene, the higher its association score to the phenotype (\nameref{sfig:consensus_stats}B). To leverage on their strengths and compensate their respective weaknesses, we built a consensus subnetwork that captures the mechanisms most shared among the solution subnetworks (Section~\ref{methods:methods}). This subnetwork (Fig~\ref{fig:consensus}) contains 93 genes and exhibits the aforementioned properties of the individual solutions: enrichment in breast cancer susceptibility genes and higher betweenness centrality than the rest of the genes.

\begin{figure}[!ht]
  \centering
  \includegraphics[width=.7\linewidth]{./figures/figure_3.pdf}
  \caption{ {\bf Consensus subnetwork on GENESIS (Section~\ref{methods:methods}) }. Each node is represented by a pie chart, which shows the methods that selected it. We labeled (and enlarged) the two most central genes (\emph{COPS5} and \emph{OFD1}) and those genes that are known breast cancer susceptibility genes and/or significantly associated with breast cancer susceptibility in the BCAC dataset. The latter ones are also colored in pink. All gene names are indicated in \nameref{sfig:consensus-names}.}
  \label{fig:consensus}
\end{figure}

Topologically the consensus consists of a connected component composed of 49 genes, and multiple smaller subnetworks. Among the latter, 19 genes are in subnetworks containing a single gene or two connected nodes, implying that they do not have a consistently altered neighborhood, but are strongly associated themselves and hence picked by at least two methods. The opposite would be the case of highly central genes in the PPIN, a property which is weakly anti-correlated with the P\=/value of association to the disease (Pearson correlation coefficient~=~-0.26, \nameref{sfig:consensus_stats}D). This suggests that they were selected because they were on the shortest path between two highly associated genes. In view of this, we hypothesize that highly central genes might contribute to the heritability through alterations of their neighborhood, consistently with the omnigenic model of disease \cite{boyle_expanded_2017}. For instance, the most central node in the consensus network is COPS5, a component of the COP9 signalosome which regulates multiple signalling pathways. \emph{COPS5} is related to multiple hallmarks of cancer and is overexpressed in multiple tumors, including breast and ovarian cancer \cite{liu_jab1_cops5_2018}. Despite its lack of association in GENESIS or BCAC (VEGAS2 P\=/value of 0.22 and 0.14 respectively), its neighbors in the consensus subnetwork have consistently low P\=/values (median VEGAS2 P\=/value~=~0.006).

We compared the results of different network methods to the European sample of the Breast Cancer Association Consortium (BCAC) \cite{Michailidou2017}, the largest GWAS to date on breast cancer (Section~\ref{methods:bcac}). Although BCAC case-control studies do not necessarily target cases with a family history of breast cancer, this comparison is pertinent since we expect a shared genetic architecture at the gene level, at which most network metods operate. This shared genetic architecture, together with BCAC's scale (90 times more samples than GENESIS) provides a reasonable counterfactual of what we would expect if GENESIS had a larger sample size. We computed a gene association score on BCAC, in an equivalent way to the one described in Section~\ref{methods:node_score}. The solutions provided by the different network approaches overlap significantly with BCAC findings (Fisher's exact test P\=/value \textless~0.019). The gene-based network methods achieve comparable precision (2\%-25\%) and recall (1.3-12.1\%) at recovering BCAC-significant genes (Fig~\ref{fig:solution_overview}C). Interestingly, while SConES GI at the SNP-level achieves a similar recall (8.6\%), it shows a much higher precision (47.3\%).

We compared the six network methods in a 5-fold subsampling setting (Section \ref{methods:comparison}). Specifically, we measured five properties (Fig~\ref{fig:benchmark}): size of the solution subnetwork; sensitivity and specificity of an L1-penalized logistic regression classifier on the selected SNPs; stability; and computational runtime. The solution size varies greatly between the different methods (Fig~\ref{fig:benchmark}A). Heinz produced the smallest solutions, with an average of 182 selected SNPs. The largest solutions came from SConES GI (6\,256.6 SNPs), and dmGWAS (4\,255.0 SNPs). To determine whether the selected SNPs could be used for patient classification, we computed the performance of the classifier on the \emph{test dataset} (Fig~\ref{fig:benchmark}B). The different classifiers displayed similarly poor sensitivities and specificities, all in the 0.52 -- 0.56 range. Interestingly, the classifier trained on all the SNPs had a similar performance, despite being the only method aiming only at minimizing prediction error. It should be considered that, although these performances are low, we do not expect to separate cases from controls well using exclusively genetic data.

\begin{figure}[!ht]
  \centering
  \includegraphics[width=\linewidth]{./figures/figure_4.pdf}
  \caption{{ \bf Comparison of network-based GWAS methods on GENESIS. } Each method was run 5 times on a random subset containing 80\% of the samples, and tested on the remaining samples (Section \ref{methods:comparison}). As LEAN did not select any gene, it was excluded from all panels except \textbf{D}. \textbf{(A)} Number of SNPs selected by each method and number of SNPs in the active set found by the classifier. Points are the average over the 5 runs; lines represent the standard error of the mean. A grey diagonal line with slope 1 is added for comparison. For reference, the active set of Lasso using all the SNPs included, on average, 154\,117.4 SNPs. \textbf{(B)} Sensitivity and specificity on test set of the L1-penalized logistic regression trained on the features selected by each of the methods. In addition, the performance of the classifier trained on all SNPs is displayed. Points are the average over the 5 runs; lines represent the standard error of the mean. \textbf{(C)} Pairwise Pearson correlations of the solutions used by different methods. A Pearson correlation of 1 means the two solutions are the same. A Pearson correlation of 0 means that there is no SNP in common between the two solutions. \textbf{(D)} Runtime of the evaluated methods, by type of network used (gene or SNP). For gene network-based methods, inverted triangles represent the runtime of the algorithm itself, and circles the total time, which includes the algorithm themselves and the additional 119\,980 seconds (1 day and 9.33 hours) that VEGAS2 took on average to compute the gene scores from SNP summary statistics.}
  \label{fig:benchmark}
  \end{figure}

Another desirable quality of an algorithm is the stability of the solution with regards to small changes in the input (Section~\ref{methods:algorithm_comparison}). Both heinz and LEAN displayed a high stability in our benchmark, consistently selecting the same genes and no genes over the 5 subsamples, respectively (Fig~\ref{fig:benchmark}C). Conversely, the other methods displayed similarly low stabilities.

In terms of computational runtime, the fastest method was heinz (Fig~\ref{fig:benchmark}D), which returned a solution in a few seconds. HotNet2 was the slowest (3 days and 14 hours on average). Including the time required to compute the gene scores, however, slows down considerably gene-based methods; on this benchmark, that step took on average 1 day and 9.33 hours. Considering that, it took 5 days on average for HotNet2 to produce a result.

In this dataset, heinz's solution is very conservative, providing a small solution with the lowest median P\=/value for the subnetwork (Table~\ref{tab:gene_solutions}). Due to this parsimonious and highly associated solution, it was the best method to stably select a set of biomarkers (Fig~\ref{fig:benchmark}C). Its conservativeness stems from its preprocessing step, which models the gene P\=/values as a mixture model of a beta distribution and a uniform distribution, controlled by an FDR parameter. Due to the limited signal at the gene level in this dataset (\nameref{sfig:snp_gene_manhattan}B), only 36 of all the genes retain a positive score after that transformation. Yet, this small solution does not provide much insight into the susceptibility mechanisms to cancer. Importantly, it ignores genes that are associated to cancer in this dataset like \emph{FGFR2}.

On the other end of the spectrum, dmGWAS, HotNet2, and SigMod produced large solutions. dmGWAS' subnetwork is the least associated subnetwork on average. This is due to the greedy framework it uses, which has a bias for larger solutions \cite{nikolayeva_network_2018}. It considered all nodes at distance 2 of the examined subnetwork, and accepted a weakly associated genes if it was linked to another, strongly associated one. This is exacerbated when the results of successive greedy searches are aggregated, leading to a large, tightly connected cluster of unassociated genes (Fig~\ref{fig:issues}A). This relatively low signal-to-noise ratio combined with the large solution requires additional analyses to draw conclusions, such as enrichment analyses. In the same line, HotNet2's subnetwork is even harder to interpret, being composed of 440 genes divided into 135 subnetworks. Lastly, SigMod misses some of the most strongly associated, breast cancer susceptibility genes in the dataset, like \emph{FGFR2} and \emph{TOX3}.

\begin{figure}[!ht]
  \centering
  \includegraphics[width=\linewidth]{./figures/figure_2.pdf}
  \caption{{\bf Drawbacks encountered when using network guided methods.} \textbf{(A)}~DmGWAS solution subnetwork. Genes with a P\=/value < 0.1 are highlighted in red. \textbf{(B)}~Centrality degree and -log\textsubscript{10} of the VEGAS2 P\=/value in BCAC for each of the nodes in the PPIN. We highlighted the genes selected by each method, and the ones selected by more than one (``Consensus''). We labeled the three most central genes that were picked by any method. \textbf{(C)}~Overlap between the solutions of SConES GS, GM or GI in the different genomic regions. SNPs that were not selected in the studied network, but were selected in another one, are displayed in background color. }
  \label{fig:issues}
  \end{figure}

Another peculiarity of network methods is their relationship to degree centrality. On the one hand we observed that highly central genes often had no association to disease (Fig~\ref{fig:issues}B). On the other, network methods favor central genes, as they often connect high scoring nodes. This was specially the case of SigMod, which selected three highly central, unassociated genes: \emph{COPS5}, \emph{CUL3} and \emph{FN1}. As we showed in Section \ref{results:consensus}, and will show in \ref{results:stable-consensus}, there is evidence in the literature of the contribution of the first two to breast cancer susceptibility. With regards to \emph{FN1}, it encodes a fibronectin, a protein of the extracellular matrix involved in cell adhesion and migration. Overexpression of \emph{FN1} has been observed in breast cancer \cite{Ioachim2002}, and it is negatively correlated with poor prognosis in other cancer types \cite{Yi2016,Sponziello2016}.
